Concurrent Programming in Harmony

### 2021-12-24 book/paper.tex
Updated chapter 4

\begin{figure}
\begin{code}
\begin{verbatim}
   0 Frame __init__ ()
code/Up.hny:1 count = 0
   1 Push 0
   2 Store count
code/Up.hny:2 done = [ False, False ]
   3 Push [False, False]
   4 Store done
code/Up.hny:4 def incrementer(self):
   5 Jump 32
   6 Frame incrementer self
code/Up.hny:5     count = count + 1
   7 Load count
   8 Push 1
   9 2-ary +
   10 Store count
code/Up.hny:6     done[self] = True
   11 Push ?done
   12 LoadVar self
   13 Address
   14 Push True
   15 Store
code/Up.hny:7     await done[1 - self]
   16 Push 1
   17 LoadVar self
   18 2-ary -
   19 Load done
   20 Apply
   21 JumpCond False 16
code/Up.hny:8     assert count == 2
   22 ReadonlyInc
   23 AtomicInc
   24 Load count
   25 Push 2
   26 2-ary ==
   27 Load count
   28 Assert2
   29 AtomicDec
   30 ReadonlyDec
   31 Return
\end{verbatim}
\end{code}
\caption{The first part of the HVM bytecode corresponding to \autoref{fig:inc}}
\label{fig:inccode}
\end{figure}

@@

Compiling the code in \autoref{fig:inc} results in the HVM bytecode
listed in \autoref{fig:inccode}.
You can obtain this code by invoking \texttt{harmony} with the \texttt{-a} flag
like so:
\begin{code}
\begin{verbatim}
harmony -a Up.hny
\end{verbatim}
\end{code}
Each thread in the HVM is predominantly a \emph{stack machine},
\index{stack machine}%
but it also has two registers.
Like shared memory, the registers usually contain dictionaries so they
can represent the values of multiple named variables.
As mentioned before, one register contains a dictionary with the local
variables of the method that the thread is currently executing,
while the other contains thread-local state.

\index{sequential consistency}%
All instructions are atomically executed.
The Harmony memory model is \emph{sequentially consistent}: all
accesses are in program order.
Most instructions pop values from the stack or push values onto the stack.
At first there is one thread, named \texttt{\_\_init\_\_},
which initializes the state.
It starts executing at instruction 0 and keeps executing until
it reaches the last instruction in the program.
In this case, it executes instructions 0 through 5 first.
The last instruction in that sequence is a \texttt{JUMP}
instruction that sets the program counter to 32
(skipping over the code for <{incrementer}> method).
The \texttt{\_\_init\_\_} thread then executes instructions
32 through 40 and finishes.
Once initialization completes, any threads that were spawned
(in this case <{incrementer(0)}> and \texttt{incrementer(1)})
can run.

\begin{figure}
\begin{code}
\begin{verbatim}
#states 32
Safety Violation
T0: __init__() [0-5,32-40] { count: 0, done: [ False, False ] }
T1: incrementer(0) [ 6- 9] { count: 0, done: [ False, False ] }
T2: incrementer(1) [ 6-18] { count: 1, done: [ False, True  ] }
T1: incrementer(0) [10-28] { count: 1, done: [ True,  True  ] }
Harmony assertion failed: 1
open file:///Users/rvr/github/harmony/harmony.html for more information
\end{verbatim}
\end{code}
\caption{The text output of running Harmony on \autoref{fig:inc}}
\label{fig:incoutput}
\end{figure}

@@

Harmony can report the following failure types:
\begin{itemize}
\item \texttt{Safety violation}: This means something went wrong with
at least one of the executions of the program that it tried.  This
can include a failing assertion, behavior violations, divide by zero,
using an uninitialized
or non-existent variable, dividing a set by an integer, and so on.
% It also includes certain infinite loops.
Harmony will print a trace of the shortest bad execution that it found.
\item \texttt{Non-terminating State}: Harmony found one or more states
from which there does not exist an execution such that all threads
terminate.  Harmony will not only print the non-terminating state with
the shortest trace, but also the list of threads
at that state, along with their program counters.
\item \texttt{Behavior Violation}:  The program terminated in a state
not allowed by the behavioral specification.
\item \texttt{Active Busy Waiting}:  There are states in which
some thread cannot make progress without the help of another thread but
does not block;
\item \texttt{Data Race}: There are states in which two or more threads
concurrently access a shared variable, at least one of which is a store
operation.
% \item \texttt{Stopped States}: Similar to non-terminating states,
% these are states in which some threads are left that cannot
% terminate.  \autoref{ch:synch} will explain what it means for
% a thread to be stopped.
\end{itemize}

Harmony checks for these types of failure conditions in the given
order: if there are multiple failure conditions, only the first is
reported.
You will learn about behavioral violations in \autoref{ch:testing}.
\emph{Active busy waiting} (\autoref{ch:condwait}) is not
technically an indication of a synchronization problem, but instead an
indication of an inefficient solution to a synchronization problem---
one that uses up significantly CPU cycles.
A \emph{data race} (\autoref{ch:synch}) may not be a bug either---whether or
not it is might depend on the semantics of the underlying memory
operations and are therefore generally undesirable.
Harmony may also warn about behaviors, in particular if the generated
behavior is only a subset of the provided behavior.

Going back to our example, Harmony reports a safety violation.
In particular, it reports that the assertion failed
because <{count}> has value 1 instead of 2.
Next it reports an execution that failed this assertion.
The program got to the failed assertion in 4 ``turns.''
The output has four columns:
\begin{enumerate}
\item A thread identifier;
\item The main method and argument of the thread;
\item The sequence of program counters of the HVM instructions that the thread executed;
\item The contents of the shared memory.
\end{enumerate}

\begin{figure}
\begin{center}
\includegraphics[width=0.9\textwidth]{figures/Up1.pdf}
\end{center}
\caption{The
\harmonyref{Up.html}{HTML output of running Harmony on \autoref{fig:inc}}.
There are four sections.
The top shows the turns to the problematic state.
Each turn shows a thread identifier, the method the thread is executing,
a ``timeline'' containing a block for each instruction executed, the
program counter of the thread, and the values of the shared variables
at those turns.
The middle section shows the line of code that is being executed.
The bottom left shows the bytecode of the program.
The bottom right shows the state of each thread.
Each row in the stack trace of a thread shows
the method that is being evaluated and
the values of its method variables.}
\label{fig:inchtml1}
\end{figure}

@@

\begin{enumerate}
\item Thread \texttt{\_\_init\_\_} (with identifier T0)
sets shared variable <{count}> to 0 and
shared variable <{done}> to \texttt{[False, False]}.
\item Thread <{incrementer(0)}> (with identifier T1)
executes instructions 6 through 9, loading the value of
<{count}> but stopping just before storing 1 into <{count}>;
\item Thread \texttt{incrementer(1)} (with identifier T2)
executes instructions 6 through 18, storing 1 into
<{count}> and storing <{True}> into <{done[1]}>;
\item Thread <{incrementer(0)}>
continues execution, executing instructions 10 through 28
storing value 1 into <{count}> (instruction 10),
storing <{True}> into <{done[0]}> (instructions 11 through 15)
finding that <{done[1]}> is <{True}> (instructions 16 through 21),
and finally detecting that the assertion is violated
(instructions 22 through 28).
\end{enumerate}

@@

In the top right, the HTML file contains the reported issue in red.
Underneath it, a table shows the four turns in the execution.
Instead of listing explicitly the program counters of the executed
instructions, the HTML file contains a list of blocks for each
executed instruction.  We call this the \emph{timeline}.
You can click on such a block to see the state of the Harmony
virtual machine just after executing the corresponding instruction.
The turn that is being executed is highlighted in green.
The table also lists the program counter of the thread at each turn,
and the values of the shared variables.

Underneath the table it shows the line of Harmony code that is being
executed in blue.

>>>>

\begin{figure}
\begin{code}
\begin{verbatim}
   0 Frame __init__ ()
code/Up.hny:1 count = 0
   1 Push 0
   2 Store count
code/Up.hny:2 done = [ False, False ]
   3 Push [False, False]
   4 Store done
code/Up.hny:4 def incrementer(self):
   5 Jump 35
   6 Frame incrementer self
code/Up.hny:5     count = count + 1
   7 Load count
   8 Push 1
   9 2-ary +
   10 Store count
\end{verbatim}
\end{code}
\caption{The first part of the HVM bytecode corresponding to \autoref{fig:inc}}
\label{fig:inccode}
\end{figure}

@@

Compiling the code in \autoref{fig:inc} results in the HVM bytecode
listed in \autoref{fig:inccode}.
You can obtain this code by invoking \texttt{harmony} with the \texttt{-a} flag
like so:
\begin{code}
\begin{verbatim}
harmony -a Up.hny
\end{verbatim}
\end{code}
Each thread in the HVM is predominantly a \emph{stack machine},
\index{stack machine}%
but it also a register.
Like shared memory, the register contains a dictionary so it
can represent the values of multiple named variables.
%
\index{sequential consistency}%
All instructions are atomically executed.
The Harmony memory model is \emph{sequentially consistent}: all
accesses are in program order.
Most instructions pop values from the stack or push values onto the stack.
At first there is one thread, named \texttt{\_\_init\_\_},
which initializes the state.
It starts executing at instruction 0 and keeps executing until
it reaches the last instruction in the program.
In this case, it executes instructions 0 through 5 first.
The last instruction in that sequence is a \texttt{JUMP}
instruction that sets the program counter to 35
(skipping over the code for <{incrementer}> method).
The \texttt{\_\_init\_\_} thread then executes the
remaining instructions and finishes.
Once initialization completes, any threads that were spawned
(in this case <{incrementer(0)}> and \texttt{incrementer(1)})
can run.

\begin{figure}
\begin{code}
\begin{verbatim}
#states 44 (time 0.001+0.001)
Safety Violation
T0: __init__()     [0-5,35-43]   { count: 0, done: [ False, False ] }
T2: incrementer(1) [6-9]         { count: 0, done: [ False, False ] }
T1: incrementer(0) [6-20]        { count: 1, done: [ True,  False ] }
T2: incrementer(1) [10-24,26-31] { count: 1, done: [ True,  True  ] }
Harmony assertion failed
open code/Up.htm for more information
\end{verbatim}
\end{code}
\caption{The text output of running Harmony on \autoref{fig:inc}}
\label{fig:incoutput}
\end{figure}

@@

Harmony can report the following failure types:
\begin{itemize}
\item \texttt{Safety violation}: This means something went wrong with
at least one of the executions of the program that it tried.  This
can include a failing assertion, behavior violations, divide by zero,
using an uninitialized
or non-existent variable, dividing a set by an integer, and so on.
% It also includes certain infinite loops.
Harmony will print a trace of the shortest bad execution that it found.
\item \texttt{Non-terminating State}: Harmony found one or more states
from which there does not exist an execution such that all threads
terminate.  Harmony will not only print the non-terminating state with
the shortest trace, but also the list of threads
at that state, along with their program counters.
\item \texttt{Behavior Violation}:  The program can terminate in a state
not allowed by the behavioral specification (\autoref{ch:testing}).
\item \texttt{Active Busy Waiting}:  There are states in which
some thread cannot make progress without the help of another thread,
but does not block (\autoref{ch:condwait}).
\item \texttt{Data Race}: There are states in which two or more threads
concurrently access a shared variable, at least one of which is a store
operation (\autoref{ch:synch}).
% \item \texttt{Stopped States}: Similar to non-terminating states,
% these are states in which some threads are left that cannot
% terminate.  \autoref{ch:synch} will explain what it means for
% a thread to be stopped.
\end{itemize}

Harmony checks for these types of failure conditions in the given
order: if there are multiple failure conditions, only the first is
reported.
\emph{Active busy waiting} (\autoref{ch:condwait}) is not
technically an indication of a synchronization problem, but instead an
indication of an inefficient solution to a synchronization problem---
one that uses up CPU cycles unnecessarily.
A \emph{data race} may not be a bug either---whether or
not it is might depend on the semantics of the underlying memory
operations and are therefore generally undesirable.
Harmony may also warn about behaviors, in particular if the generated
behavior is only a subset of the provided behavior.

Going back to our example, Harmony reports a safety violation.
In particular, it reports that an assertion failed.
The program got to the failed assertion in four ``turns.''
The output has four columns:
\begin{enumerate}
\item A thread identifier;
\item The main method and argument of the thread;
\item The sequence of program counters of the HVM instructions that the thread executed;
\item The contents of the shared memory.
\end{enumerate}

\begin{figure}
\begin{center}
\includegraphics[width=0.9\textwidth]{figures/Up1.png}
\end{center}
\caption{The
\harmonyref{Up.html}{HTML output of running Harmony on \autoref{fig:inc}}.
There are four sections.
The top shows the turns to the problematic state.
Each turn shows a thread identifier, the method the thread is executing,
a ``timeline'' containing a block for each instruction executed, the
program counter of the thread, the values of the shared variables
at those turns, and any values the program may have printed.
The middle section shows the line of code that is being executed.
The bottom left shows the bytecode of the program.
The bottom right shows the state of each thread.
Each row in the stack trace of a thread shows
the method that is being evaluated and
the values of its method variables.}
\label{fig:inchtml1}
\end{figure}

@@

\begin{enumerate}
\item Thread \texttt{\_\_init\_\_} (with identifier T0)
executes instructions 0 through 5 and 35 through 43, setting
shared variable <{count}> to 0 and
shared variable <{done}> to <{[False, False]}>.
\item Thread <{incrementer(0)}> (with identifier T2)
executes instructions 6 through 9, loading the value of
<{count}> but stopping just before storing 1 into <{count}>;
\item Thread \texttt{incrementer(1)} (with identifier T1)
executes instructions 6 through 20, storing 1 into
<{count}> and storing <{True}> into <{done[1]}>;
\item Thread <{incrementer(0)}> (T2)
continues execution, executing instructions 10 through 24
storing value 1 into <{count}> (instruction 10),
storing <{True}> into <{done[0]}>,
finding that <{done[1]}> is <{True}>,
and finally detecting that the assertion is violated.
\end{enumerate}

@@

In the top right, the HTML file contains the reported issue in red.
Underneath it, a table shows the four turns in the execution.
Instead of listing explicitly the program counters of the executed
instructions, the HTML file contains a list of blocks for each
executed instruction.  We call this the \emph{timeline}.
You can click on such a block to see the state of the Harmony
virtual machine just after executing the corresponding instruction.
The turn that is being executed is highlighted in green.
The table also lists the program counter of the thread at each turn,
the values of the shared variables, and any values the thread
may have printed (none in this case).
%
Underneath the table it shows the line of Harmony code that is being
executed in blue.
